### 2013-01-20 branches/CUDA/BookGPU/Chapters/chapter14/ch14.tex
submitten version
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1872 40876441-3496-0410-a613-a2fca35b2ae2

\section{Introduction}
The lattice Boltzmann (LB) method (for an overview see, e.g.,
\cite{succi-book}) has become a popular approach to a variety of fluid
dynamics problems.  It provides a way to solve the incompressible,
isothermal Navier-Stokes equations and has the attractive features of
being both explicit in time and local in space. This makes the LB
method well-suited to parallel computation. Many efficient parallel
implementations of the LB method have been undertaken, typically using
a combination of distributed domain decomposition and the Message
Passing Interface (MPI). However, the potential
performance benefits offered by GPUs has motivated a new `mixed-mode'
approach to address very large problems. Here, fine-grained
parallelism is implemented on the GPU, while MPI is reserved for
larger-scale parallelism.  This mixed mode is of increasing interest
to application programmers at a time when many supercomputing services
are moving
toward clusters of GPU accelerated nodes. The design questions which
arise when developing a lattice Boltzmann code for this type of
heterogeneous system are therefore worth studying. Furthermore, similar
questions also recur in many other types of stencil-based algorithms.

@@

\section{Background}\label{ch14:sec:background}

@@

\begin{figure}[!t]
\centering
\includegraphics[width=12cm]{Chapters/chapter14/figures/decomphalo}
\caption{Left: the lattice is decomposed between MPI tasks. For
  clarity we show a 2D decomposition of a 3D lattice, but in practice
  we decompose in all 3 dimensions. Halo cells are added to each
  sub-domain (as shown on the upper right for a single slice) which store
  data retrieved from remote neighbours in the halo exchange. Lower
  right: the D3Q19 velocity set resident on a lattice site;
  highlighted are the 5 ``outgoing'' elements to be transferred in a
  specific direction.}
\label{ch14:fig:decomphalo}
\end{figure}

@@

\section{Single GPU Implementation}\label{ch14:sec:singlegpu}

@@

\section{Multiple GPU Implementation}\label{ch14:sec:parallelgpu}

@@

\begin{figure}[!t]
\centering
\includegraphics[width=10cm]{Chapters/chapter14/figures/two_graphs_ud}
\caption{The weak (top) and strong (bottom) scaling of \textit{Ludwig}.  Closed
  shapes denote results using the CPU version run on the Cray
  XE6 (using two 16-core AMD Interlagos CPUs per node), while open
  shapes denote results using the GPU version on the Cray XK6 (using a
  single NVIDIA X2090 GPU per node). Top: the benchmark time is shown
  where the problem size per node is constant. Bottom: performance is
  shown where, for each of the four problem sizes presented, the
  results are scaled by the lattice volume, and all results are
  normalized by the same arbitrary constant.  }
\label{ch14:fig:scaling}
\end{figure}

@@

\section{Moving Solid Particles}\label{ch14:sec:particles}

\begin{figure}[t]
\centering
%\includegraphics[width=10cm]{Chapters/chapter14/figures/bbl}
\includegraphics[width=10cm]{Chapters/chapter14/figures/colloid_new}
\caption{
A two-dimensional schematic picture of spherical particles on the lattice.
Left: a particle is allowed
to move continuously across the lattice, and the position of the
surface defines fluid lattice sites (light blue) and solid lattice
sites (dark red). The discrete surface is defined by links
where propagation would intersect the surface (arrows). Note the
discrete shape of the two particles is different. Right: post-collision
distributions are reversed at the surface by the process of bounce-back
on links, which replaces the propagation.
%A 2D illustration of colloidal particles (grey circles)
%  interacting with the lattice, where each lattice point is
%  represented by a cross. Left: each particle spans multiple sites.
%  Right: the particle-facing distribution components are moved inside
%  the particle with the direction reversed, such that the ``bounce
%  back'' will be completed when the fluid propagates.
}
\label{ch14:fig:bbl}
\end{figure}

>>>>

\section{Introduction} \index{lattice Boltzmann} \index{complex fluids}
The lattice Boltzmann (LB) method (for an overview see, e.g.,
\cite{succi-book}) has become a popular approach to a variety of fluid
dynamics problems.  It provides a way to solve the incompressible,
isothermal Navier-Stokes equations and has the attractive features of
being both explicit in time and local in space. This makes the LB
method well-suited to parallel computation. Many efficient parallel
implementations of the LB method have been undertaken, typically using
a combination of distributed domain decomposition and the Message
Passing Interface (MPI). However, the potential
performance benefits offered by GPUs has motivated a new `mixed-mode'
approach to address very large problems. Here, fine-grained
parallelism is implemented on the GPU, while MPI is reserved for
larger-scale parallelism.  This mixed mode is of increasing interest
to application programmers at a time when many supercomputing services
are moving
toward clusters of GPU accelerated nodes. The design questions which
arise when developing a lattice Boltzmann code for this type of
heterogeneous system are therefore worth studying. Furthermore, similar
questions also recur in many other types of stencil-based algorithms.

@@

\section{Background}\label{ch14:sec:background} \index{lattice Boltzmann} \index{complex fluids} \index{MPI}

@@

\begin{figure}[!t]
\centering
\includegraphics[width=12cm]{Chapters/chapter14/figures/decomphalo}
\caption[Parallel decomposition of D3Q19 lattice.]{Left: the lattice is
  decomposed between MPI tasks. For clarity we show a 2D decomposition
  of a 3D lattice, but in practice we decompose in all 3 dimensions.
  Halo cells are added to each sub-domain (as shown on the upper right
  for a single slice) which store data retrieved from remote
  neighbours in the halo exchange. Lower right: the D3Q19 velocity set
  resident on a lattice site; highlighted are the 5 ``outgoing''
  elements to be transferred in a specific direction.}
\label{ch14:fig:decomphalo}
\end{figure}

@@

\section{Single GPU Implementation}\label{ch14:sec:singlegpu} \index{CUDA}

@@

\section{Multiple GPU Implementation}\label{ch14:sec:parallelgpu} \index{multi-GPU programming} \index{MPI} \index{parallel scaling}

@@

\begin{figure}[!t]
\centering
\includegraphics[width=10cm]{Chapters/chapter14/figures/two_graphs_ud}
\caption[Scaling of Ludwig to many GPUs in parallel.]{The weak (top)
  and strong (bottom) scaling of \textit{Ludwig}.  Closed shapes
  denote results using the CPU version run on the Cray XE6 (using two
  16-core AMD Interlagos CPUs per node), while open shapes denote
  results using the GPU version on the Cray XK6 (using a single NVIDIA
  X2090 GPU per node). Top: the benchmark time is shown where the
  problem size per node is constant. Bottom: performance is shown
  where, for each of the four problem sizes presented, the results are
  scaled by the lattice volume, and all results are normalized by the
  same arbitrary constant.  }
\label{ch14:fig:scaling}
\end{figure}

@@

\section{Moving Solid Particles}\label{ch14:sec:particles} \index{particle-fluid interaction} \index{molecular dynamics}

\begin{figure}[t]
\centering
%\includegraphics[width=10cm]{Chapters/chapter14/figures/bbl}
\includegraphics[width=10cm]{Chapters/chapter14/figures/colloid_new}
\caption[Interaction of spherical particles with lattice Boltzmann fluid.]{
A two-dimensional schematic picture of spherical particles on the lattice.
Left: a particle is allowed
to move continuously across the lattice, and the position of the
surface defines fluid lattice sites (light blue) and solid lattice
sites (dark red). The discrete surface is defined by links
where propagation would intersect the surface (arrows). Note the
discrete shape of the two particles is different. Right: post-collision
distributions are reversed at the surface by the process of bounce-back
on links, which replaces the propagation.
%A 2D illustration of colloidal particles (grey circles)
%  interacting with the lattice, where each lattice point is
%  represented by a cross. Left: each particle spans multiple sites.
%  Right: the particle-facing distribution components are moved inside
%  the particle with the direction reversed, such that the ``bounce
%  back'' will be completed when the fluid propagates.
}
\label{ch14:fig:bbl}
\end{figure}
